Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-06-08 Paper/paper.tex
adding figures for sec 3

The process of measuring systematic and statistical measurement errors
involves simulating many artificial GW signals, and inferring source binary
parameters from them.
We start with generating a signal waveform, using the model LEA+, and injecting
it in zero noise to obtain a stretch of data $d_n$. Source binary parameters
$\vec{\Theta}:=\vec{\theta}\cup\{\lambdans\}$ are reconstructed from this
injected signal. Using Bayes' theorem, their joint inferred probability
distribution can be calculated as
\begin{equation}\label{eq:postprob}
 p(\vec{\Theta} | d_n, K) = \dfrac{p(d_n|\vec{\Theta}, K)\,p(\vec{\Theta} | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\Theta} | K)$ is the a-priori probability of binary parameters
taking particular values. Throughout this paper, we impose a prior uniform on
individual component masses, BH spin, and NS's tidal deformability. In
addition, we restrict mass-ratios to $q\geq 2$, as LEA+ is not valid for 
$1\leq q\leq 2$. $p(d_n|\vec{\Theta}, K)$ is the
likelihood of obtaining the given stretch of data $d_n$ if we assume that a
signal parameterized by $\vec{\Theta}$ is buried in it, and is given by
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\Theta}) \equiv p(d_n| \vec{\Theta}, K) = \mathcal{N}\, \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\Theta})$ is a filter template with parameters 
$\vec{\Theta}$, and $\langle\cdot|\cdot\rangle$ is a suitably defined
detector-noise weighted inner-product~\footnote{The inner product
$\langle\cdot|\cdot\rangle$  is defined as
\begin{equation}
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
\end{equation}
where $\tilde{a}(f)$ is the Fourier transform of the finite time series $a(t)$,
and $S_n(|f|)$ is the one-sided amplitude spectrum of detector noise.}
The denominator in Eq.~\ref{eq:postprob} is the a-priori probability of finding
the particular signal in $d_n$ and we assume that each injected signal is as
likely as any other. Having constructed the probability distribution function
$p(\vec{\Theta} | d_n, K)$, extracting the measured probability distribution
for a single parameter (say $\alpha$) involves integrating
\begin{equation}\label{eq:marginalize}
 p(\alpha | d_n, K) = \int\D \vec{\Theta}_\alpha\, p(\vec{\Theta} | d_n, K),
\end{equation}
where $\vec{\Theta}_\alpha$ is the set of remaining parameters, i.e.
$\vec{\Theta}_\alpha:=\vec{\Theta} - \{\alpha\}$.

@@

One simplifying assumption made to mitigate 
computational cost was setting the frequency sampling rate $\Delta f=0.4$~Hz, which
we found to be sufficient to calculate likelihoods robustly~\cite{Purrer:2015nkh}.
% \prayush{TOCheck: MICHAEL}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{How is PE affected by not including NS matter effects?}\label{s1:PEwithnoNS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

of NS-BH sources. We do so by comparing the systematic biases introduced by this 
assumption to the overall statistical uncertainties in the measurement of binary 
mass and spin parameters. We find that unless the single-detector SNR of the signal
exceeds $30$ (for design Advanced LIGO sensitivity), the bias accurued by ignoring
tidal effects in filtering templates will remain smaller than the statistical 
uncertainty of the same measurement.

We first take a set of NSBH binary waveforms, which include the modeling of spin and
tidal effects. Binary mass-ratio is sampled
uniformly over the range $q\in[2,5]$ taking $q=\{2,3,4,5\}$ as our test points. The
mass of the NS is fixed to $\mns=1.35M_\odot$, and therefore knowing $q$ gives $\mbh$.
The spin of the NS is fixed to zero. The spin of the BH is sampled uniformly over the
range $\chibh\in[-0.75, +0.75]$, choosing $\chibh=\{-0.5, 0, +0.5, +0.75\}$ as our
test points. Finally, we vary the deformability parameter uniformly over 
$\lambdans\in[0,2000]$, choosing $\lambdans=\{500, 1000, 1500, 2000\}$ plus an additional
$\lambdans=800$ case. Waveform are generated for all possible combinations of these
parameters, and rescaled to different strengths with $\rho=\{20, 30, 50, 70\}$. For all
of the resulting signals, we perform a fully Bayesian parameter estimation aimed at
measuring binary masses and spins from the signal. Our analysis
templates differ from the signal in only one sense: all tidal terms are set to zero
in the former, i.e. they correspond to BHBH mergers.

>>>>

The process of measuring systematic and statistical measurement errors
involves simulating many artificial GW signals, and inferring source binary
parameters from them using Bayesian statistics.
We start with generating a signal waveform, using the model LEA+, and injecting
it in zero noise to obtain a stretch of data $d_n$. Source binary parameters
$\vec{\Theta}:=\vec{\theta}\cup\{\lambdans\}$ are reconstructed from this
injected signal. Using Bayes' theorem, their joint inferred probability
distribution of $\vec{\Theta}$ can be written as
\begin{equation}\label{eq:postprob}
 p(\vec{\Theta} | d_n, K) = \dfrac{p(d_n|\vec{\Theta}, K)\,p(\vec{\Theta} | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\Theta} | K)$ is the a-priori probability of binary parameters
taking particular values. Throughout this paper, we impose a prior uniform on
individual component masses, BH spin, and NS's tidal deformability. In
addition, we restrict mass-ratios to $q\geq 2$, as LEA+ is not valid for 
$1\leq q\leq 2$. $p(d_n|\vec{\Theta}, K)$ is the
likelihood of obtaining the given stretch of data $d_n$ if we assume that a
signal parameterized by $\vec{\Theta}$ is buried in it, and is given by
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\Theta}) \equiv p(d_n| \vec{\Theta}, K) = \mathcal{N}\, \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\Theta})$ is a filter template with parameters 
$\vec{\Theta}$, and $\langle\cdot|\cdot\rangle$ is a suitably defined
detector-noise weighted inner-product~\footnote{The inner product
$\langle\cdot|\cdot\rangle$  is defined as
\begin{equation}
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
\end{equation}
where $\tilde{a}(f)$ is the Fourier transform of the finite time series $a(t)$,
and $S_n(|f|)$ is the one-sided amplitude spectrum of detector noise.}
The denominator in Eq.~\ref{eq:postprob} is the a-priori probability of finding
the particular signal in $d_n$ and we assume that each injected signal is as
likely as any other. Having constructed the probability distribution function
$p(\vec{\Theta} | d_n, K)$, extracting the measured probability distribution
for a single parameter (say $\alpha$) involves integrating
\begin{equation}\label{eq:marginalize}
 p(\alpha | d_n, K) = \int\D \vec{\Theta}_\alpha\, p(\vec{\Theta} | d_n, K),
\end{equation}
where $\vec{\Theta}_\alpha$ is the set of remaining parameters, i.e.
$\vec{\Theta}_\alpha:=\vec{\Theta} - \{\alpha\}$.

@@

One simplification we make to mitigate computational cost is to set the
frequency sampling interval to $\Delta f=0.4$~Hz, which we find to be
sufficient for robust likelihoods calculations~\cite{Purrer:2015nkh}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{How is PE affected if we ignore NS matter effects?}\label{s1:PEwithnoNS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

of NS-BH sources.
% We do so by comparing the systematic biases introduced by this 
% assumption to the overall statistical uncertainties in the measurement of binary 
% mass and spin parameters. We find that unless the single-detector SNR of the signal
% exceeds $30$ (for design Advanced LIGO sensitivity), the bias accurued by ignoring
% tidal effects in filtering templates will remain smaller than the statistical 
% uncertainty of the same measurement.

\red{Reorder:
We construct NSBH waveforms with LEA+. We inject them, and run MCMC on the injections
using BBH templates. We perform this analysis for [... parameters ...]. We always
use $\mns=1.35M_\odot$, but explore $\lambdans=\{500,1000,1500,2000\}$.
}

\prayush{
We first take a set of NSBH binary waveforms, which include the modeling of spin and
tidal effects.
For all of the resulting signals, we perform a fully Bayesian parameter estimation
aimed at measuring binary masses and spins from the signal. Our analysis
templates differ from the signal in only one sense: all tidal terms are set to zero
in the former, i.e. they correspond to BHBH mergers. Binary mass-ratio is sampled
uniformly over the range $q\in[2,5]$ taking $q=\{2,3,4,5\}$ as our test points. The
mass of the NS is fixed to $\mns=1.35M_\odot$, and therefore knowing $q$ gives $\mbh$.
The spin of the NS is fixed to zero. The spin of the BH is sampled uniformly over the
range $\chibh\in[-0.75, +0.75]$, choosing $\chibh=\{-0.5, 0, +0.5, +0.75\}$ as our
test points. Finally, we vary the deformability parameter uniformly over 
$\lambdans\in[0,2000]$, choosing $\lambdans=\{500, 1000, 1500, 2000\}$ plus an additional
$\lambdans=800$ case. Waveform are generated for all possible combinations of these
parameters, and rescaled to different strengths with $\rho=\{20, 30, 50, 70\}$.
}
